---
title: "Understanding_voom"
author: "Lucas Boatwright"
date: "5/12/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Voom

This script was created to more fully understand how limma voom functions.

```{r data}
library("edgeR")
library("airway")

# Load data
data(airway)
d <- DGEList(assay(airway))
samples <- colnames(d)
dim(d)

# Filter at a minimum of 10 Counts-per-million
keep <- rowSums(cpm(d) > cpm(10,mean(d$samples$lib.size))[1]) == 8
d <- d[keep,]

# Generate design matrix
trt = factor(c(1,2,1,2,1,2,1,2), labels=c("untreated","treated"))
design=model.matrix(~0 + trt)
```

Below, I've taken the basic portions of the voom function and explained what's happening in each command. Portions of the function were removed and the remaining code functions with the assumption that the design matrix is appropriate, expression matrix is a DGElist and there is no normalization.

```{r voom_simplified}
span = 0.5 # Function default
# Get library size
lib.size <- colSums(d$counts)
# Generate output list
out <- list()
# Get gene names from DGE list
out$genes <- d$genes
# Get sample names from DGE list
out$targets <- d$samples
# Get log2 CPM
y <- t(log2(t(d$counts + 0.5)/(lib.size + 1) * 1e+06))
# Fit log2-CPM ~ design
# This is a genewise linear model fit to expression data using least squares
fit <- lmFit(y, design)
# Get fit mean as log2 counts -- converted back from log2-CPM
sx <- fit$Amean + mean(log2(lib.size + 1)) - log2(1e+06)
# Get fit std dev -- relative to log2-CPM
sy <- sqrt(fit$sigma)

# Fit lowess
l <- lowess(sx, sy, f = span)
# l contain both x and y coordinates for lowess fit
# approxfun -- Returns a list of points which linearly interpolate given data points
# rule 2, the value at the closest data extreme is used
f <- approxfun(l, rule = 2)

# If the number of groups in the fit is less than the number of columns
if (fit$rank < ncol(design)) {
  # If less, get those pivot values
  j <- fit$pivot[1:fit$rank]
  # Duplicate the coefficienct columns pivot-times
  fitted.values <- fit$coef[, j, drop = FALSE] %*% t(fit$design[, 
                                                                j, drop = FALSE])
} else {
  # Duplicate the coefficient columns according to the design matrix columns
  # Which is expressed as log2-CPM
  fitted.values <- fit$coef %*% t(fit$design)
}
# Convert from log2-CPM to CPM
fitted.cpm <- 2^fitted.values
# Convert from CPM to counts
fitted.count <- 1e-06 * t(t(fitted.cpm) * (lib.size + 1))
# Convert from counts to log2-counts
fitted.logcount <- log2(fitted.count)
# Calculate inverse weights using the linear interpolation generated from the lowess fit on the log2-counts
# A weight is inverse variance not Sqrt(STDEV), so it's to the fourth power
w <- 1/f(fitted.logcount)^4
# This is probably a check to make sure the dimensions are the same...
dim(w) <- dim(fitted.logcount)
# Output expression is log2-CPM
out$E <- y
# Weights are inverse variance
out$weights <- w
# Design matrix
out$design <- design

# Output targets are sample names with associated library sizes
if (is.null(out$targets)) {
  out$targets <- data.frame(lib.size = lib.size)
} else {out$targets$lib.size <- lib.size}

# I moved the plot to the bottom of the function since it's self-explanatory after going through the function
# Plot voom mean-variance trend -- assuming voom(d, plot=TRUE)
plot = TRUE
if (plot) {
  # sx = log2 counts, sy = Sqrt( log2-CPM STDEV )
  plot(sx, sy, xlab = "log2( count size + 0.5 )", ylab = "Sqrt( standard deviation )", 
       pch = 16, cex = 0.25)
  title("voom: Mean-variance trend")
  lines(l, col = "red")
}



```

```{r session_info}
sessionInfo()
```